Astrophysical calibration of gravitational-wave detectors

### 2015-09-22 docs/cosmic_cal.tex
Added text to the standard siren and GRB sections plus refs.

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}
\acrodef{MCMC}[MCMC]{Markov chain Monte Carlo}

@@

We should explain what this means starting with reference
to~\cite{1986Natur.323..310S}.

@@

% what are sGRBs 
\cm{What are sGRBs?}\\

>>>>

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{NS}[NS]{neutron star}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}
\acrodef{PSD}[PSD]{power spectral density}
\acrodef{MCMC}[MCMC]{Markov chain Monte Carlo}
\acrodef{ET}[ET]{Einstein Telescope}

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type 1a Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the detector network. However, measurement of the phase evolution of such an
event allows accurate determination of the chirp mass~\footnote{In reality it
is the redshifted chirp mass that is measured.}. Timing information from the
different signal arrival times at each detector in the network allows
reasonably accurate sky position determination.  Finally, amplitude variation
between differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.  A great help to this final
stage for this project is the assumption that the source has an \ac{sGRB}
counterpart.  It follows that we therefore have a well constrained prior belief
on the inclination of the source orbit with respect to the line-of-sight based
on the expected level of beaming in the \ac{sGRB} emission.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosty distance to sources.
This is distinct from Type 1a Supernovae standard candles that only provide
relative luminosty distance measures and require calibration via other methods
as part of the cosmological distance ladder. At first sight the missing
component for standard sirens is a complementary measurement of the source
redshift which would allow each event to inform the distance-redshift
relationship and hence obtain cosmological parameter estimates.  However, it
was not thought possible (until recently) that redshift measurements could be
obtained from \ac{GW} observations. In~\cite{DelPozzo:2011uf} (building on the
original idea proposed in~\cite{1986Natur.323..310S} the idea of identifying
host galaxies within the sky position uncertainty region was developed. Each
galaxy would provide a redshift measure and would contribute to a disparate mix
of possible cosmological parameters. However, by combining multiple (${\sim}20$)
detections, the true set of cosmological parameters emerges with an expected
${\sim}5\%$ uncertainty on the Hubble constant.

% continued use of sirens
The use of \ac{GW} events with \ac{sGRB} counterparts was investigated
in~\cite{2010CQGra..27u5006S} and~\cite{2011PhRvD..83b3005Z} with respect to
the 3rd generation \ac{GW} interferometer, the \ac{ET}. It was
assumed in both studies that 1000 coincident events would be observed and
cosmological parameter estimation ability was compared with current \ac{EM}
results with focus on dark energy parameters. Other methods for cosmological
inference (see~\cite{1996PhRvD..53.2878F},~\cite{2012PhRvD..85b3535T},
and~\cite{2012PhRvD..86b3502T}) have been proposed for \ac{BNS} systems that do
not rely on any redshift measurements and instead use the distribution of
measured \acp{SNR} combined with some assumptions regarding the \ac{NS} mass
and spatial distributions. Finally we mention that for \ac{BNS} systems it has
been shown in~\cite{Messenger:2011ux} and~\cite{2013arXiv1312.1862M} that features
of the tidal and post-merger hyper-massive \ac{NS} stages of the \ac{GW}
waveform can allow the measurement of the source redshift if the \ac{NS}
equation-of-state is sufficiently well known. This development, and seperately, the method
based on \ac{SNR} distributions will allow \ac{GW} cosmological inference
without the need for complementary \ac{EM} observations.

% put this in context with calibration
To put the existing work on \ac{GW} cosmology in context with the study
described in this manuscript we are essentially turning the standard
cosmological problem upside-down. We assume that the existing \ac{EM}
cosmological parameter estimation (specifically for the Hubble constant) is
accurately determined and therefore a known quantity. We then assume that the
unknown in our analysis is the absolute amplitude calibration of the
interferometers in the global \ac{GW} detector network. The issue of
calibration has largely been ignored in \ac{GW} cosmological inference studies
and in Sec.~\ref{sec:discussion} we discuss some of the implications of
combining these issues in future analyses.

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\acp{BNS}~\cite{1989Natur.340..126E},~\cite{1992ApJ...395L..83N}. The emmission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of \ac{BNS} mergers. The distances
to these objects are obtained via identification of

The observed redshift range is from about 0.2 to 2 for short GRBs (sGRBs), with
a mean of about 0.4. The XRT commonly observes sGRBs up to z ≃ 0.5 but in 10.5
yr of operation Swift has not seen one with a measured redshift z < 0.1, i.e.,
the NS-NS merger aLIGO target GW horizon ∼400 Mpc, suggesting that such events
may be quite rare.
Estimates for jet beaming are θj ∼ 5◦ for lGRBs and θj ∼ 5 − 15◦ for sGRBs
(Burrows et al. 2006, Grupe et al. 2006, Fong et al. 2012). Beaming angles for
sGRBs are still highly uncertain. The beaming factors fb = 1 − cos θj ≃ θj2/2
are roughly 1/300 for lGRBs and 1/30 for sGRBs. Based on the observed rate of
sGRBs by Swift, Coward et al. (2012) estimate a LVC detection rateof∼3−30yr−1
forθj ≃ 15◦. Chen&Holz(2013)claim3−7yr−1 forGRBGW+EM detections. Kelley et al.
(2013) estimate the rate of Swift or Fermi observations joint with LVC
detections to be ∼0.07 yr−1. Siellez, Boe ̈r, & Gendre (2014) consider current
and future high energy missions and estimate a rate of simultaneous GW+EM
detections of ∼0.1 − 4 yr−1 in the LVC era. Wanderman & Piran (2015) estimate a
co-detection rate LVC+Fermi of 0.1 − 1 yr−1 and LVC+Swift of 0.02 − 0.14 yr−1.
